---
title: "PLIER pathways analysis"
output: html_notebook
---

```{r}
meaningful_difference <- 0.2
```

### What additional pathways are identified by PLIER after doubling the sample size?

Here we look for cancer pathways that are found more frequently in the data with the full sample size than in data with half the sample size.
We ran PLIER 10 times using each combination of data normalization method and sample size (half or full).
Out of those 10 runs, we found the proportion of times each pathway was significantly associated with a latent variable.
We set an arbitrary "meaningful difference" threshold of `r meaningful_difference` to detect when that proportion was meaningfully greater using full sample size compared to half sample size.
We also arbitrarily limit results to those pathways found in over half of runs using the full size data.

### Load packages and pathways

```{r}
library(tidyverse)
library(PLIER)

data(bloodCellMarkersIRISDMAP)
data(canonicalPathways)
data(oncogenicPathways)
data(svmMarkers)
```

### Read in data for each cancer type
```{r}
brca_pathways_df <- read_tsv("plots/data/BRCA_subtype_PLIER_pathways.tsv") %>%
  mutate(cancer_type = "BRCA")
gbm_pathways_df <- read_tsv("plots/data/GBM_subtype_PLIER_pathways.tsv") %>%
  mutate(cancer_type = "GBM")
```

### Filter data to identify pathways detected more frequently in full data
```{r}

bind_rows(brca_pathways_df, gbm_pathways_df) %>%
  filter(FDR < 0.05) %>%
  group_by(nmeth, pseq, pathway, cancer_type) %>%
  summarize(n_seeds = length(unique(seed_index)),
            prop_seeds = n_seeds/10,
            .groups = "drop") %>%
  mutate(nmeth = str_replace(nmeth, " ", "_")) %>%
  mutate(nmeth = str_replace(nmeth, "-", "_")) %>%
  mutate(nmeth = str_remove_all(nmeth, "\\(")) %>%
  mutate(nmeth = str_remove_all(nmeth, "\\)")) %>%
  mutate(nmeth_pseq = str_c(nmeth, pseq, sep = "_")) %>%
  select(cancer_type, pathway, prop_seeds, nmeth_pseq) %>%
  pivot_wider(id_cols = c("cancer_type", "pathway"),
              names_from = nmeth_pseq,
              values_from = prop_seeds) %>%
  mutate_if(is.numeric, replace_na, 0) %>%
  filter(log_0 - array_only_50 >= .2,
         log_100 - seq_only_50 >= .2,
         log_0 > 0.5,
         log_100 > 0.5,
         npn_50 > 0.5) %>%
  select(cancer_type, pathway, array_only_50, seq_only_50, log_0, log_100, npn_50) %>%
  filter(pathway %in% colnames(oncogenicPathways)) %>%
  arrange(cancer_type) %>%
  knitr::kable()
```

### Citation:
Mao W, Zaslavsky E, Hartmann BM, Sealfon SC, Chikina M. Pathway-level information extractor (PLIER) for gene expression data. Nat Methods. 2019;16: 607–610.